---
title: 'Data Imputation: MICE'
author: 'Mingcheng Hu'
format: 
    pdf:
        toc: true
        keep-tex: true
        include-in-header: 
            text: |
                \usepackage{fvextra}
                \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
                \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r}
library(mice)
library(ggplot2)
library(dplyr)
library(corrplot)
options(scipen = 999)  # disable scientific notation

setwd("/work/users/y/u/yuukias/BIOS-Material/BIOS992/src/step3_impute_split_data/impute_data")
```


# Import and Preprocess Data

```{r}
data_unimputed <- read.csv("full_data.csv")
(dim(data_unimputed))
head(data_unimputed)
```

**For the missing rate, check `export_data.ipynb`.**

Then, we should declare the type of categorical variables

```{r}
data_unimputed$ethnicity <- as.factor(data_unimputed$ethnicity)
data_unimputed$sex <- as.factor(data_unimputed$sex)
data_unimputed$smoking <- as.factor(data_unimputed$smoking)
data_unimputed$diabetes <- as.factor(data_unimputed$diabetes)
data_unimputed$hypertension_treatment <- as.factor(data_unimputed$hypertension_treatment)
data_unimputed$education <- as.factor(data_unimputed$education)
data_unimputed$activity <- as.factor(data_unimputed$activity)
```

We can also visualize the correlation matrix of the numeric variables.

```{r, fig.width=12, fig.height=10}
data_numeric <- data_unimputed[, sapply(data_unimputed, is.numeric)]
# exclude eid
data_numeric <- subset(data_numeric, select = -eid)
cor_matrix <- cor(data_numeric, use = "pairwise.complete.obs")
corrplot(cor_matrix, 
         method = "color",      
         type = "upper",       
         order = "original",      
         addCoef.col = "black", 
         tl.col = "black",    
         tl.srt = 45,          
         diag = FALSE,
         na.label = "NA")
```

There are no strong correlations among the numeric variables. Only the cholesterol variables and ECG-related variables are moderately correlated.


# MICE Imputation

## Run Imputation

We only need one imputed dataset for subsequent analysis.

```{r, label = "MICE Imputation", cache = TRUE}
start_time <- proc.time()
tryCatch({
    data_imputed <- mice(
        data_unimputed, 
        m = 1, maxit = 50, seed = 1234, print=FALSE)
}, error = function(e) {
    print(e)
    return(NULL)
})
end_time <- proc.time()
run_time <- end_time - start_time
print(run_time)

save(data_imputed, file = "full_data_imputed_mice.RData")
```


## Check Imputation Results


```{r}
plot(data_imputed)  # note it is not dataframe, but an object
```

```{r}
stripplot(data_imputed, pch = 20, cex = 1.2) 
```

```{r}
densityplot(data_imputed)
```

