---
title: 'Data Imputation: MissForest'
author: 'Mingcheng Hu'
format: 
    pdf:
        toc: true
        keep-tex: true
        include-in-header: 
            text: |
                \usepackage{fvextra}
                \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
                \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r}
library(missForest)
library(ggplot2)
library(dplyr)
library(corrplot)
library(parallel)
library(doParallel)
options(scipen = 999)  # disable scientific notation

setwd("/work/users/y/u/yuukias/BIOS-Material/BIOS992/src/step3_impute_split_data/impute_data")
```

# Import and Preprocess Data

**We will impute the data based on `eligible_data.csv`.**

```{r}
data_unimputed <- read.csv("eligible_data.csv")
(dim(data_unimputed))
head(data_unimputed)
```

**For the missing rate, check `export_data.ipynb`.**

Then, we should declare the type of categorical variables

```{r}
data_unimputed$ethnicity <- as.factor(data_unimputed$ethnicity)
data_unimputed$sex <- as.factor(data_unimputed$sex)
data_unimputed$smoking <- as.factor(data_unimputed$smoking)
data_unimputed$diabetes <- as.factor(data_unimputed$diabetes)
data_unimputed$hypertension_treatment <- as.factor(data_unimputed$hypertension_treatment)
data_unimputed$education <- as.factor(data_unimputed$education)
data_unimputed$activity <- as.factor(data_unimputed$activity)
```

We can also visualize the correlation matrix of the numeric variables.

```{r, fig.width=12, fig.height=10}
data_numeric <- data_unimputed[, sapply(data_unimputed, is.numeric)]
# exclude eid
data_numeric <- subset(data_numeric, select = -eid)
cor_matrix <- cor(data_numeric, use = "pairwise.complete.obs")
corrplot(cor_matrix, 
         method = "color",      
         type = "upper",       
         order = "original",      
         addCoef.col = "black", 
         tl.col = "black",    
         tl.srt = 45,          
         diag = FALSE,
         na.label = "NA")
```

There are no strong correlations among the numeric variables. Only the cholesterol variables and ECG-related variables are moderately correlated.

# MissForest Imputation

## Run Imputation

```{r, label = "MissForest Imputation", cache = TRUE}
start_time <- proc.time()
tryCatch({
    n_cores <- min(detectCores() - 1, 8)
    registerDoParallel(cores = n_cores)

    data_imputed <- missForest(data_unimputed, 
        ntree = 100, maxiter = 10, verbose = TRUE,
        parallel = "variables")

    stopImplicitCluster()
}, error = function(e) {
    print(e)
    return(NULL)
})
end_time <- proc.time()
run_time <- end_time - start_time
print(run_time)

save(data_imputed, file = "eligible_data_imputed_missForest.RData")
```

## Check Imputation Results

We can plot densities of both the observed and imputed values of all variables to see whether the imputations are reasonable. Differences in the densities between the observed and imputed values may suggest a problem that needs to be further checked.

```{r}
par(mfrow = c(1, 2))
for(col in c("hdl_chol", "total_chol")) {
    plot(density(data_unimputed[[col]], na.rm = TRUE), 
            main = col,
            col = "blue", lwd = 2)
    lines(density(data_imputed$ximp[[col]]), 
            col = "red", lwd = 2)
    legend("topright", 
            legend = c("original", "imputed"), 
            col = c("blue", "red"), 
            lwd = 2)
}
```


Fortunately, there is no significant difference between the densities of the original and imputed values for our case.