---
title: 'Evaluate Model Performance'
author: 'Mingcheng Hu'
params:
    adjust_type:
        label: 'Adjustment Type of Survival Model'
        value: 'full'
        choices:
            - 'minimal'
            - 'partial'
            - 'full'
    impute_type:
        label: 'Imputation Type'
        value: 'imputed'
        choices:
            - 'unimputed'
            - 'imputed'
    include_statin:
        label: 'Include Participants Taking Statin in the Model'
        value: 'no'
        choices:
            - 'yes'
            - 'no'
format: 
    pdf:
        toc: true
        keep-tex: true
        include-in-header: 
            text: |
                \usepackage{fvextra}
                \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
                \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r output = FALSE}
library(tidyverse)
library(survival)
library(survcomp) # general way to calculate concordance index
library(glmnet)
library(randomForestSRC)
library(xgboost)
library("SHAPforxgboost")
library(kableExtra) # include knitr automatically

source("/work/users/y/u/yuukias/BIOS-Material/BIOS992/utils/csv_utils.r")
# * Don't use setwd() for Quarto documents!
# setwd("/work/users/y/u/yuukias/BIOS-Material/BIOS992/data")

adjust_type <- ifelse(exists("params"), params$adjust_type, "full") # options: "minimal", "partial", "full"
impute_type <- ifelse(exists("params"), params$impute_type, "imputed") # options: "unimputed", "imputed"
include_statin <- ifelse(exists("params"), params$include_statin, "no") # options: "yes", "no"

set.seed(1234)
```

```{r}
# string of parameters
adjust_type_str <- switch(adjust_type,
    minimal = "minimal",
    partial = "partial",
    full = "full"
)
print(paste0("Model Adjustment Type: ", adjust_type_str))
impute_type_str <- switch(impute_type,
    unimputed = "unimputed",
    imputed = "imputed"
)
print(paste0("Data Imputation Type: ", impute_type_str))
```

# Load Models

```{r}
load(get_data_path("cox_model_univariate", adjust_type_str, impute_type_str, include_statin, model = "cox"))
load(get_data_path("cox_model_multivariate", adjust_type_str, impute_type_str, include_statin, model = "cox"))
load(get_data_path("cox_model_lasso", adjust_type_str, impute_type_str, include_statin, model = "cox"))
load(get_data_path("cox_model_step", adjust_type_str, impute_type_str, include_statin, model = "cox"))

load(get_data_path("rsf_model", adjust_type_str, impute_type_str, include_statin, model = "rsf"))

load(get_data_path("xgb_var_select_name", adjust_type_str, impute_type_str, include_statin, model = "xgb"))
load(get_data_path("xgb_model", adjust_type_str, impute_type_str, include_statin, model = "xgb"))
```


# Load Data

```{r}
if (include_statin == "yes") {
    data_test <- read.csv(paste0("/work/users/y/u/yuukias/BIOS-Material/BIOS992/data/test_data_", impute_type_str, "_statin.csv"),
        header = TRUE
    )
} else {
    data_test <- read.csv(paste0("/work/users/y/u/yuukias/BIOS-Material/BIOS992/data/test_data_", impute_type_str, ".csv"),
        header = TRUE
    )
}

data_test <- data_test[, -1] # the first column is the index generated by sklearn
(dim(data_test))
```

```{r}
data <- select_subset(data_test, type = adjust_type)
(dim(data))
```

```{r}
colnames(data)
```

```{r}
data <- tibble::as_tibble(data)
```

```{r label = "Full Data Specific Processing"}
# * There are some imputed ethnicity set to "e". We will exclude them at this time.
data <- data %>%
    filter(ethnicity != "e")

# * We also need to manually relevel the categorical variables
data <- data %>%
    mutate(
        # Set "Never" (0) as baseline for smoking
        smoking = factor(smoking,
            levels = c("0", "1", "2", "-3"),
            labels = c("Never", "Previous", "Current", "Prefer not to answer")
        ),
        
        # Set "No" (0) as baseline for diabetes
        diabetes = factor(diabetes,
            levels = c("0", "1", "-1", "-3"),
            labels = c("No", "Yes", "Do not know", "Prefer not to answer")
        ),
        
        # Ensure other categorical variables are properly factored
        ethnicity = factor(ethnicity,
            levels = c("1", "2", "3", "4", "5", "6"),
            labels = c("White", "Mixed", "Asian/Asian British", "Black/Black British", "Chinese", "Other")
        ),
        
        education = factor(education,
            levels = c("1", "2", "3", "4", "5", "6", "-7", "-3"),
            labels = c("College/University degree", "A levels/AS levels",
                      "O levels/GCSEs", "CSEs", "NVQ/HND/HNC",
                      "Other professional", "None of the above",
                      "Prefer not to answer")
        ),
        
        activity = factor(activity,
            levels = c("0", "1", "2"),
            labels = c("Low", "Moderate", "High")
        ),
        
        sex = factor(sex,
            levels = c("0", "1"),
            labels = c("Female", "Male")
        ),
        
        hypertension_treatment = factor(hypertension_treatment,
            levels = c("0", "1"),
            labels = c("No", "Yes")
        )
    )
```

```{r}
# * It is very hard to compare the HR as different predictors are on different magnitudes, so we need to normalize them.
time_col <- data$time
event_col <- data$event
data <- data %>%
    select(-c(time, event)) %>%
    mutate(across(where(is.numeric), scale)) %>%
    mutate(
        time = time_col,
        event = event_col
    )
```

**Note now the interpretation of HR is different! For example, if HR=1.16 for the predictor in the univariate model fitted using scaled data, it means that each standard deviation increase is associated with 16% higher risk of event.**

```{r label = "Model-Specific Data Preparation"}
# For Cox model:
data_complete <- na.omit(data)

# For RSF model: We don't need to exclude the missing values

# For XGBoost model:
test_x <- data_complete %>% select(-c(time, event))
test_x_xgb <- model.frame(~ . - 1, data = test_x, na.action = na.pass)
test_x_xgb <- model.matrix(~ . - 1, data = test_x_xgb)
test_y_lower_bound <- data_complete$time
test_y_upper_bound <- ifelse(data_complete$event == 1, data_complete$time, Inf)
dtest <- xgb.DMatrix(
    data = test_x_xgb,
    label_lower_bound = test_y_lower_bound,
    label_upper_bound = test_y_upper_bound
)
dtest_selected <- xgb.DMatrix(
    data = test_x_xgb[, vars_selected],
    label_lower_bound = test_y_lower_bound,
    label_upper_bound = test_y_upper_bound
)
```

# Model Performance Evaluation and Comparison

## Cox Models

```{r}
# Multivariate Cox Model (only on complete data for fair comparison)
concord_full1 <- concordance(cox_model_full_complete, newdata = data_complete)
print(paste0("Concordance of Multivariate Cox Model: ", round(concord_full1$concordance, 3)))
```

```{r}
pred_full <- predict(cox_model_full_complete, newdata = data_complete, type = "risk")
concord_full2 <- concordance.index(pred_full, data_complete$time, data_complete$event)$c.index
lower_full <- concordance.index(pred_full, data_complete$time, data_complete$event)$lower
upper_full <- concordance.index(pred_full, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of Multivariate Cox Model: ", round(concord_full2, 3), " (", round(lower_full, 3), ", ", round(upper_full, 3), ")"))
```

Both approaches give the same result.

```{r}
# LASSO
# * Now we should use Cindex() instead of concordance()
x_test <- model.matrix(~ . - 1 - time - event, data = data_complete)
y_test <- Surv(data_complete$time, data_complete$event)
pred_lasso <- predict(cox_model_lasso, newx = x_test, s = "lambda.1se")
concord_lasso1 <- apply(pred_lasso, 2, Cindex, y = y_test)
print(paste0("Concordance of LASSO Cox Model: ", round(concord_lasso1, 3)))
```

```{r}
pred_lasso2 <- predict(cox_model_lasso, newx = x_test, s = "lambda.1se")
concord_lasso2 <- concordance.index(pred_lasso2, data_complete$time, data_complete$event)$c.index
lower_lasso <- concordance.index(pred_lasso2, data_complete$time, data_complete$event)$lower
upper_lasso <- concordance.index(pred_lasso2, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of LASSO Cox Model: ", round(concord_lasso2, 3), " (", round(lower_lasso, 3), ", ", round(upper_lasso, 3), ")"))
```

Both approaches give the same result.


```{r}
# Stepwise
concord_step1 <- concordance(cox_model_step, newdata = data_complete)
print(paste0("Concordance of Stepwise Cox Model: ", round(concord_step1$concordance, 3)))
```

```{r}
pred_step <- predict(cox_model_step, newdata = data_complete, type = "risk")
concord_step2 <- concordance.index(pred_step, data_complete$time, data_complete$event)$c.index
lower_step <- concordance.index(pred_step, data_complete$time, data_complete$event)$lower
upper_step <- concordance.index(pred_step, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of Stepwise Cox Model: ", round(concord_step2, 3), " (", round(lower_step, 3), ", ", round(upper_step, 3), ")"))
```

Both approaches give the same result.

## RSF Model

```{r}
data_complete_rsf <- data_complete %>% select(-c(time, event))
data_complete_rsf <- model.frame(~ . - 1, data = data_complete_rsf, na.action = na.pass)
data_complete_rsf <- model.matrix(~ . - 1, data = data_complete_rsf)
data_complete_rsf <- as.data.frame(data_complete_rsf)
```

The model with selected variables:

```{r}
pred_rsf <- predict(rsf_model, newdata = data_complete_rsf)
concord_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$c.index
lower_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$lower
upper_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of RSF Model: ", round(concord_rsf, 3), " (", round(lower_rsf, 3), ", ", round(upper_rsf, 3), ")"))
```

The model with all variables:

```{r}
pred_rsf <- predict(rsf_model_full, newdata = data_complete_rsf)
concord_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$c.index
lower_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$lower
upper_rsf <- concordance.index(pred_rsf$predicted, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of RSF Model (Full): ", round(concord_rsf, 3), " (", round(lower_rsf, 3), ", ", round(upper_rsf, 3), ")"))
```

We will also get variable importance from the RSF model.

```{r label = "Variable Importance of RSF Model", eval = FALSE}
vimp_rsf <- vimp(rsf_model)
```

```{r label = "Variable Importance of RSF Model (Full)", eval = FALSE}
vimp_rsf_full <- vimp(rsf_model_full)
```

```{r echo = FALSE, eval = FALSE}
save(vimp_rsf, 
     file = get_data_path("vimp_rsf", adjust_type_str, impute_type_str, include_statin, model = "rsf"))
save(vimp_rsf_full, 
     file = get_data_path("vimp_rsf_full", adjust_type_str, impute_type_str, include_statin, model = "rsf"))
```

## XGBoost Model

The model with selected variables:

```{r}
# Note this is the survival time, larger value means lower risk!
pred_xgb <- predict(xgb_model, dtest_selected) 
pred_xgb <- -pred_xgb # now larger value means higher risk
concord_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$c.index
lower_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$lower
upper_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of XGBoost Model: ", round(concord_xgb, 3), " (", round(lower_xgb, 3), ", ", round(upper_xgb, 3), ")"))
```


The model with all variables:

```{r}
# * Note this is the survival time, larger value means lower risk!
pred_xgb <- predict(xgb_model_full, dtest) 
pred_xgb <- -pred_xgb # now larger value means higher risk
concord_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$c.index
lower_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$lower
upper_xgb <- concordance.index(pred_xgb, data_complete$time, data_complete$event)$upper
print(paste0("Concordance of XGBoost Model (Full): ", round(concord_xgb, 3), " (", round(lower_xgb, 3), ", ", round(upper_xgb, 3), ")"))
```
